---
title: "Spotify and the Russo-Ukrainian War"
subtitle: "Many openly pro- and anti-Kremlin Russian artists are on Spotify. How has their popularity changed since the start of the war in Ukraine, and "
author: "Ruty Korotaev"
thanks: "Code and data are available at: https://github.com/rutykorotaev/finalpaper"
date: "19 April 2023"
abstract: "Following Russia's escalated invasion of Ukraine in February 2022, countries across the West and around the world have sought to end business dealings with Russian companies and boycott pro-Russian public figures. This widespread effort has gone from the highest levesl of business and government, to individuals opting to stop purchasing Russian goods and not engaging with Russian popular culture, including musicians. This report uses data collected from Spotify APIs to address two main research questions. First, how has the popularity and following of both pro- and anti-Kremlin Russian musicians changed since the onset of the war in Ukraine. And two, how are Spotify's recommendation features impacting engagement with these Russian artists, and is it able to differentiate between the political affiliations of these artists? Among the main findings is______. "
format: pdf
number-sections: true
bibliography: references.bib
---

# Introduction

Amid much of the Western world's efforts to sanction Kremlin-affiliated enterprises and Russian businesses, the movement to limit engagement with Russian popular culture has been an interesting side effect of the conflict. Following Russia's escalated invasion of Ukraine in February 2022, many Ukrainians and Ukrainian allies have rejected any forms of cultural engagement with Russian artists and the Russian language, which ranges from boycotting Western businesses still operating in the Russia and has trickled down to the music people listen to. This research paper looks at data collected through Spotify API to understand how the popularity of openly pro- and anti- Kremlin Russian musicians has shifted since the war began, looking at a time period between February and October 2022, taking into account the six months during which there was the most action in terms of sanctioning and banning Russian outlets (and vice versa within Russia and its bans of Western organizations). Among the pro-Kremlin artists are ____, while the anti-Kremlin artists - name of whom have now been exiled out of the country - are ___. The paper will also explore the relationship between an artist's popularity rating and their follower count, to see how it may have shifted and changed over time. 

Further, the second part of this paper will take from literature around other social media outlets which analyzes how recommendations can lead users down certain rabbit holes. While these can sometimes be harmless, the worst cases can lead people down pathes to radicalization, entrenching users into information environments that spread mis- and disinformation, conspiracy theories, and far-right rhethoric. This paper analyzes how Spotify's recommendation features encourages or discourages users to engage with Russian users, and whether it is able to distinguish between the political affiliations of different artists. 

This paper is divided into several key sections. The Data section {@sec-data}, will outline where the data is from and what main variables were explored in the dataset. All of the data visualizations will be included in this section and will seek to demonstrate how the popularity of each performer has evolved over time. Next, the Results section {@sec-results} will provide an overview of the main findings that can be made by the data visualizations, which will be further explored within the {@sec-discussion} section, which will have three main discussion points that draw on conclusions that can be made based on the data. The first point {@sec-first-point} describes how____, while the second point {@sec-second-point} explains. Finally, the third point {@sec-third-point} stipulates that ____. The project findings and limitations can be found in the last section {@sec-limitations}, which will discuss the various issues with the dataset and its findings, as well as potential changes that could be made for future projects. 

# Data {#sec-data}

This paper uses data collected using the Spotify API "spotifyr" [@citespotifyr], and visualizes data from five top Russian artists and 5 top Ukrainian artists to understand how and if their popularity on Spotify has been affected since the start of the escalated invasion of Ukraine. As part of this, the main variables that were used in this project include popularity, year, artist name, and track number. This paper uses R [@citeR] to analyze the dataset, and several R packages were included in the project, including "tidyverse" [@citetidyverse], "dplyr" [@citedplyr], "kableExtra" [@citekableExtra], and "knitr" [@citeknitr].

## Popularity: Pro- vs. Anti-Kremlin Russian Artists on Spotify

```{r}
#| echo: false
#| message: false
#| label: fig-popularity
#| fig-cap: Popularity of Anti and Pro-Kremlin Russian Artists on Spotify Amid the War in Ukraine

# Load  libraries and read in dataset 
library(tidyverse)
library(ggplot2)

clean_info_dataset <- 
  read_csv(here::here("inputs/data/clean_info_dataset.csv"))

# Create data frame with all the necessary artist names
artists <- c("oxxxymiron", "noizemc", "face", "icepeak", "maxkorzh", "skriptonite", "basta", "eldzhey")

# Extract data from the popularity column from the clean_info_dataset
popularity <- clean_info_dataset$popularity

# Create a data frame with the artist names and their popularity and followers.total values
data <- data.frame(x = artists, y = popularity)

# Plot the data in a line segment, with the pro-Kremlin artists being red and the anti-Kremlin artists being in blue
ggplot(data, aes(x = artists, y = popularity)) + 
  geom_segment(aes(x = artists, xend = artists, y = 0, yend = y),
color = ifelse(data$x %in% c("oxxxymiron", "noizemc", "face", "icepeak"), "blue", "red"),
linewidth = ifelse(data$x %in% c("oxxxymiron", "noizemc", "face", "icepeak"), 0.3, 0.3)) +
  geom_point(color = ifelse(data$x %in% c("oxxxymiron", "noizemc", "face", "icepeak"), "blue", "red"), 
linewidth = ifelse(data$x %in% c("oxxxymiron", "noizemc", "face", "icepeak"), 2.5, 2.5))

```


```{r}
#| echo: false
#| message: false
#| label: fig-pop_followers
#| fig-cap: Correlation Between Follower Count and Popularity Score

library(ggrepel)

# Create data frame of artists 
artists_pop_followers <- c("oxxxymiron",
             "noizemc",
             "face",
             "icepeak",
             "maxkorzh",
             "skriptonite",
             "basta",
             "eldzhey")

# Extract all the popularity data for each artist
oxxxymiron_pop <- oxxxymiron$popularity
noizemc_pop <- noizemc$popularity
face_pop <- face$popularity
icepeak_pop <- icepeak$popularity
maxkorzh_pop <- maxkorzh$popularity
skriptonite_pop <- skriptonite$popularity
basta_pop <- basta$popularity
eldzhey_pop <- eldzhey$popularity

popularity <- c(oxxxymiron_pop,
                noizemc_pop,
                face_pop,
                icepeak_pop,
                maxkorzh_pop,
                skriptonite_pop,
                basta_pop,
                eldzhey_pop)

# Extract all the follower data for each artist
oxxxymiron_followers <- oxxxymiron$followers.total
noizemc_followers <- noizemc$followers.total
face_followers <- face$followers.total
icepeak_followers <- icepeak$followers.total
maxkorzh_followers <- maxkorzh$followers.total
skriptonite_followers <- skriptonite$followers.total
basta_followers <- basta$followers.total
eldzhey_followers <- eldzhey$followers.total

followers <- c(oxxxymiron_followers,
               noizemc_followers,
               face_followers,
               icepeak_followers,
               maxkorzh_followers,
               skriptonite_followers,
               basta_followers,
               eldzhey_followers)

scatterplot_data <- data.frame(x = popularity, y = followers)

#Plot graph 
ggplot(scatterplot_data, aes(x = popularity, y = followers)) +
geom_point(aes(colour = factor(artists_pop_followers)), size = 1.75)

```


# Model {#sec-model}

```{r}
#| echo: false
#| message: false
#| label: fig-skriptonite_recs
#| fig-cap: Spotify's Recommended Artists for Skriptonite 

lm(y ~ x, data)

# Create a dataframe
data <- data.frame(
  option = c(1:20),  
# Assign a numeric ID to each option
  recommended = c(1,1,1,1,0,1,1,1,0,1,1,1,1,1,1,1,1,1,0,1)
)
# There should be 20 options in the recommended variable - replace all the 0's for 1's if they are pro-Russia

# Fit a logistic regression model
model <- glm(recommended ~ option, data = data, family = binomial())

# Extract the model coefficients
coefficients <- coef(model)


```


```{r}
#| echo: false
#| message: false
#| label: fig-face_recs
#| fig-cap: Spotify's Recommended Artists for FACE 


# Create a dataframe
data <- data.frame(
  option = c(1:20),  
# Assign a numeric ID to each option
  recommended = c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)
)
# There should be 20 options in the recommended variable - replace all the 0's for 1's if they are pro-Russia

# Fit a logistic regression model
model <- glm(recommended ~ option, data = data, family = binomial())

# Extract the model coefficients
coefficients <- coef(model)


```

# Results {#sec-results}

# Discussion {#sec-discussion}

## First Finding:  {#sec-first-point}
The first finding that was 

## Second Finding: {#sec-second-point}

## Third Finding: {#sec-third-point}

## Project Limitations and Next Steps {#sec-limitations}

\newpage

# References


