---
title: "Spotify and the Russo-Ukrainian War"
subtitle: "An Analysis of Russian Artist Popularity Amid the War in Ukraine"
author: "Ruty Korotaev"
thanks: "Code and data are available at: https://github.com/rutykorotaev/finalpaper"
date: "19 April 2023"
abstract: "Following Russia's escalated invasion of Ukraine in February 2022, countries across the West and around the world have sought to end business dealings with Russian companies and boycott pro-Russian public figures. This widespread effort has gone from the highest levesl of business and government, to individuals opting to stop purchasing Russian goods and not engaging with Russian popular culture, including musicians. This report uses data collected from Spotify APIs to address two main research questions. First, how has the popularity and following of both pro- and anti-Kremlin Russian musicians changed since the onset of the war in Ukraine. And two, how are Spotify's recommendation features impacting engagement with these Russian artists, and is it able to differentiate between the political affiliations of these artists? Among the main findings is______. "
format: pdf
number-sections: true
bibliography: references.bib
---

# Introduction

Amid much of the Western world's efforts to sanction Kremlin-affiliated enterprises and Russian businesses, the movement to limit engagement with Russian popular culture has been an interesting side effect of the conflict. Following Russia's escalated invasion of Ukraine in February 2022, many Ukrainians and Ukrainian allies have rejected any forms of cultural engagement with Russian artists and the Russian language, which ranges from boycotting Western businesses still operating in the Russia and has trickled down to the music people listen to. This research paper looks at data collected through Spotify API to understand how the popularity of openly pro- and anti- Kremlin Russian musicians has shifted since the war began, looking at a time period between February and October 2022, taking into account the six months during which there was the most action in terms of sanctioning and banning Russian outlets (and vice versa within Russia and its bans of Western organizations). Among the pro-Kremlin artists are ____, while the anti-Kremlin artists - name of whom have now been exiled out of the country - are ___. The paper will also explore the relationship between an artist's popularity rating and their follower count, to see how it may have shifted and changed over time. 

Further, the second part of this paper will take from literature around other social media outlets which analyzes how recommendations can lead users down certain rabbit holes. While these can sometimes be harmless, the worst cases can lead people down pathes to radicalization, entrenching users into information environments that spread mis- and disinformation, conspiracy theories, and far-right rhethoric. This paper analyzes how Spotify's recommendation features encourages or discourages users to engage with Russian users, and whether it is able to distinguish between the political affiliations of different artists. 

This paper is divided into several key sections. The Data section {@sec-data}, will outline where the data is from and what main variables were explored in the dataset. All of the data visualizations will be included in this section and will seek to demonstrate how the popularity of each performer has evolved over time. Next, the Results section {@sec-results} will provide an overview of the main findings that can be made by the data visualizations, which will be further explored within the {@sec-discussion} section, which will have three main discussion points that draw on conclusions that can be made based on the data. The first point {@sec-first-point} describes how____, while the second point {@sec-second-point} explains. Finally, the third point {@sec-third-point} stipulates that ____. The project findings and limitations can be found in the last section {@sec-limitations}, which will discuss the various issues with the dataset and its findings, as well as potential changes that could be made for future projects. 

# Data {#sec-data}

This paper uses data collected using the Spotify API "spotifyr" [@citespotifyr], and visualizes data from five top Russian artists and 5 top Ukrainian artists to understand how and if their popularity on Spotify has been affected since the start of the escalated invasion of Ukraine. As part of this, the main variables that were used in this project include popularity, year, artist name, and track number. This paper uses R [@citeR] to analyze the dataset, and several R packages were included in the project, including "tidyverse" [@citetidyverse], "dplyr" [@citedplyr], and "knitr" [@citeknitr].

## Popularity: Pro- vs. Anti-Kremlin Russian Artists on Spotify

```{r}
#| echo: false
#| message: false
#| label: fig-popularity
#| fig-cap: Popularity of Anti and Pro-Kremlin Russian Artists on Spotify Amid the War in Ukraine

# Load  libraries and read in dataset 
library(tidyverse)
library(ggplot2)
library(readr)

clean_info_dataset <- 
  read_csv(here::here("inputs/data/clean_info_dataset.csv"))

# Create data frame with all the necessary artist names
artists <- c("Oxxxymiron","Noize MC","FACE","IC3PEAK","Max Korzh","Skryptonite","Basta","Eldzhey")

# Extract data from the popularity column from the clean_info_dataset
popularity <- clean_info_dataset$popularity

# Create a data frame with the artist names and their popularity and followers.total values
data <- data.frame(x = artists, y = popularity)

# Plot the data in a line segment, with the pro-Kremlin artists being red and the anti-Kremlin artists being in blue
ggplot(data, aes(x = x, y = popularity)) + 
  geom_segment(aes(x = x, xend = artists, y = 0, yend = y),
color = ifelse(data$x %in% c("Oxxxymiron", "Noize MC", "FACE", "IC3PEAK"), "blue", "red"),
linewidth = ifelse(data$x %in% c("Oxxxymiron", "Noize MC", "FACE", "IC3PEAK"), 0.3, 0.3)) +
  geom_point(color = ifelse(data$x %in% c("Oxxxymiron", "Noize MC", "FACE", "IC3PEAK"), "blue", "red"), 
linewidth = ifelse(data$x %in% c("Oxxxymiron", "Noize MC", "Noize MC", "IC3PEAK"), 2.5, 2.5))

```

The following chart demontrates the correlation between follower count and popularity rating .


```{r}
#| echo: false
#| message: false
#| label: fig-pop_followers
#| fig-cap: Correlation Between Follower Count and Popularity Score

library(ggrepel)
library(readr)
library(dplyr)

clean_info_dataset <- 
  read_csv(here::here("inputs/data/clean_info_dataset.csv"))

# Create data frame of artists 
artists_pop_followers <- c("Oxxxymiron",
             "Noize MC",
             "FACE",
             "Ice Peak",
             "Max Korzh",
             "Skriptonite",
             "Basta",
             "Eldzhey")

# Extract all the popularity data for each artist
oxxxymiron_pop <- clean_info_dataset %>%
  filter(name == "Oxxxymiron") %>%
  pull(popularity)
noizemc_pop <- clean_info_dataset %>%
  filter(name == "Noize MC") %>%
  pull(popularity)
face_pop <- clean_info_dataset %>%
  filter(name == "FACE") %>%
  pull(popularity)
icepeak_pop <- clean_info_dataset %>%
  filter(name == "IC3PEAK") %>%
  pull(popularity)
maxkorzh_pop <- clean_info_dataset %>%
  filter(name == "Max Korzh") %>%
  pull(popularity)
skriptonite_pop <- clean_info_dataset %>%
  filter(name == "Skryptonite") %>%
  pull(popularity)
basta_pop <- clean_info_dataset %>%
  filter(name == "Basta") %>%
  pull(popularity)
eldzhey_pop <- clean_info_dataset %>%
  filter(name == "Eldzhey") %>%
  pull(popularity)

popularity2 <- c(oxxxymiron_pop,
                noizemc_pop,
                face_pop,
                icepeak_pop,
                maxkorzh_pop,
                skriptonite_pop,
                basta_pop,
                eldzhey_pop)

# Extract all the follower data for each artist
oxxxymiron_followers <- clean_info_dataset %>%
  filter(name == "Oxxxymiron") %>%
  pull(followers.total)
noizemc_followers <- clean_info_dataset %>%
  filter(name == "Noize MC") %>%
  pull(followers.total)
face_followers <- clean_info_dataset %>%
  filter(name == "FACE") %>%
  pull(followers.total)
icepeak_followers <- clean_info_dataset %>%
  filter(name == "IC3PEAK") %>%
  pull(followers.total)
maxkorzh_followers <- clean_info_dataset %>%
  filter(name == "Max Korzh") %>%
  pull(followers.total)
skriptonite_followers <- clean_info_dataset %>%
  filter(name == "Skryptonite") %>%
  pull(followers.total)
basta_followers <- clean_info_dataset %>%
  filter(name == "Basta") %>%
  pull(followers.total)
eldzhey_followers <- clean_info_dataset %>%
  filter(name == "Eldzhey") %>%
  pull(followers.total)

followers <- c(oxxxymiron_followers,
               noizemc_followers,
               face_followers,
               icepeak_followers,
               maxkorzh_followers,
               skriptonite_followers,
               basta_followers,
               eldzhey_followers)

scatterplot_data <- data.frame(x = followers, y = popularity2) 


#Plot graph 
ggplot(scatterplot_data, aes(x = popularity2, y = followers)) +
geom_point(aes(colour = factor(artists_pop_followers)), size = 1.75)

```



# Model {#sec-model}

```{r}
#| echo: false
#| message: false
#| label: fig-skriptonite_recs
#| fig-cap: Spotify's Recommended Artists for Skriptonite and FACE

library(tidyverse)
library(stargazer)

# Create a dataframe
data <- data.frame(
  Russia = c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1),  
  # Assign a numeric ID to each option
  recommendations = c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,0,1,1,1,0,1,1,1,1,1,1,1,1,1,0)
)
# There should be 20 options in the recommended variable - replace all the 0's for 1's if they are pro-Russia
#the glm model works like this - your first variable is the outcome (y) 
#and the second one, after the ~ is the predictor variable (x).
#so in this code you're looking at the influence of whether or not an artist is pro-Russia (0 = they're not)
#on whether or not the recommended artists are also pro-Russian (0 = they're not)

# Fit a logistic regression model

model <- glm(recommendations ~ Russia, data = data, family = "binomial")

# Extract the model coefficients
coefficients <- coef(model)

#figuring out the odds ratio
exp(coef(model))
#this gives you the log odds 

stargazer(model, type = "html", out = "results.htm")

```


```{r}
#| echo: false
#| message: false
#| label: fig-face_recs
#| fig-cap: Spotify's Recommended Artists for FACE 

```

# Results {#sec-results}

There are several key results that can be gleaned from the above visualizations. First, as {@fig-popularity} shows, we can see that the pro-Kremlin artists (who are outlined in red) and the anti-Kremlin artists (in blue) do not have huge differences in popularity among Spotify users. We can see that Skryptonite, the Russian rapper is among the most openly supportive of the Putin regime, ranks as the most popular, though by a very small margin. The least popular artist appears to be Oxxxymiron, who is 


# Discussion {#sec-discussion}

## First Finding:  {#sec-first-point}
The first finding, as outlined in {@fig-popularity} is that the political affilation of a certain artist does not seem to impact their popularity among listeners. 

## Second Finding: {#sec-second-point}

## Third Finding: {#sec-third-point}

## Project Limitations and Next Steps {#sec-limitations}

There are number of project limitations that should be considered when reviewing this data. First, the Spotify API posed a lot of issues when it came to accessing telling data. When initially setting out to do this project, I had intended to track the popularity of Russian and Ukrainian artists on Spotify over the course of the first six months following the Russian invasion of Ukraine. However, after many, many attempts, and becoming overly familiar with all of the functions within the Spotify R package {@citespotifyr}, it became apparent that it is impossible to access either an artist's popularity rating or monthly listeners number of the course of a specified period of time without the use of a Third-Party software to do so. Had I been successful in this, it would have been interesting to see how and if the popularity of various artists changed over the course of the war, particularly as people continued to make everyday, conscious choices in boycotting products and artists associated with Russia. Moreover, there is also a function that allows developers to see the popularity of an artist within certain Spotify markets, but it also was difficult to use and hard to actually gain this data. After a lot of trial and error, this paper was able to get a few key variables about artists, but future work in this could seek to review more deeply the question of Russsian and Ukrainian musical popularity over time and analyzing differences within various key markets. 

In addition, my own personal biases may have played a role in the decision-making process as it relates to reviewing the artist's political affiliations. As mentioned above, the criteria for deciding whether an artist was supportive of the Putin regime or not was based on a review of several factors. First, I analyzed their respective social media pages (namely Facebook, VKontakte, and Instagram) to see if they have posted any statements online about their stance on Russia's invasion of Ukraine. If an artist did not have any statements or other symbols of solidarity to Ukraine, I assumed that they were pro-Kremlin, or at the very least, unwilling to use their online platforms to speak out against the regime. This could have led to a false assumption on whether someone was supportive of Putin, because as is well-documented, saying openly critical things of the Russian government poses many personal risks to not only musicians and artists, but ordinary people. Another factor that I incorporated in this decision-making process was seeing if they had any concerts happening in Russia or if they were performing internationally. This serves as a very telling indicator because most experts say that only artists who are pro-Putin (or at least silently complicit) are still able to perform within Russia without being banned by authorities. In this sense, while I attempt to cross reference various indicators of political affilation, there is inevitably room for error in this approach. 

Finally, another important point to note is that, as mentioned above, Spotify has been banned in Russia and the organization has stopped all of its operations within the country in March 2022. While it is difficult to know how many people within Russia are using VPNs to circumvent government bans and access Spotify, this paper does not address how and if the ban has impacted Russian artist popularity since the start of the ban (largely due to the aforementioned reasons). Regardless, it is important to note that the ban must have played some kind of role in their popularity, as not all people are able to use VPNs to gain access to Spotify within Russia. Moreover, while this seems to be related to privacy issues, the Spotify API does not allow developers to access demographic data of a certain artist's listeners. This data would have provided invaluable insights, particularly in showing which countries have the most users actively listening to either pro- or anti-Kremlin artists. This data could then be tied to larger ideas on the proliferation of Russian disinformation around the world, and could offer valuable insights into the impact and permeation of pro-Kremlin narratives, and how they translate into individual's musical preferences and choices. 

\newpage
# References


